---
title: "Predicting Reading Test Score using Linear Regression: An application of Machine Learning"
author: "Huruy Belay"
date: "September 12, 2017"
output: html_document
---
Reading Test Scores

The Programme for International Student Assessment (PISA) is a test given every three years to 15-year-old students from around the world to evaluate their performance in mathematics, reading, and science. This test provides a quantitative way to compare the performance of students from different parts of the world. We will predict the reading scores of students from the United States of America on the 2009 PISA exam.

The datasets [pisa2009train.csv](https://d37djvu3ytnwxt.cloudfront.net/assets/courseware/v1/d2c0eaf1a7b1aeee047a2be54e0c6173/asset-v1:MITx+15.071x+2T2017+type@asset+block/pisa2009train.csv) and [pisa2009test.csv](https://d37djvu3ytnwxt.cloudfront.net/assets/courseware/v1/b1cf56f09e79ee9b71b38143560252be/asset-v1:MITx+15.071x+2T2017+type@asset+block/pisa2009test.csv) contain information about the demographics and schools for American students taking the exam, derived from [2009 PISA Public-Use Data Files](https://nces.ed.gov/pubsearch/pubsinfo.asp?pubid=2011038) distributed by the United States National Center for Education Statistics (NCES). While the datasets are not supposed to contain identifying information about students taking the test, by using the data you are bound by the [NCES data use agreement](https://d37djvu3ytnwxt.cloudfront.net/assets/courseware/v1/8522b75c7463c923fb96ae274fdf8213/asset-v1:MITx+15.071x+2T2017+type@asset+block/NCES_Data_Use_Agreement.txt), which prohibits any attempt to determine the identity of any student in the datasets.

Each row in the datasets[pisa2009train.csv](https://d37djvu3ytnwxt.cloudfront.net/assets/courseware/v1/d2c0eaf1a7b1aeee047a2be54e0c6173/asset-v1:MITx+15.071x+2T2017+type@asset+block/pisa2009train.csv) and [pisa2009test.csv](https://d37djvu3ytnwxt.cloudfront.net/assets/courseware/v1/b1cf56f09e79ee9b71b38143560252be/asset-v1:MITx+15.071x+2T2017+type@asset+block/pisa2009test.csv) represents one student taking the exam. The datasets have the following variables:

* **grade**: The grade in school of the student (most 15-year-olds in America are in 10th grade)

* **male**: Whether the student is male (1/0)

* **raceeth**: The race/ethnicity composite of the student

* **preschool**: Whether the student attended preschool (1/0)

* **expectBachelors**: Whether the student expects to obtain a bachelor's degree (1/0)

* **motherHS**: Whether the student's mother completed high school (1/0)

* **motherBachelors**: Whether the student's mother obtained a bachelor's degree (1/0)

* **motherWork**: Whether the student's mother has part-time or full-time work (1/0)

* **fatherHS**: Whether the student's father completed high school (1/0)

* **fatherBachelors**: Whether , (1/0)

* **fatherWork**: Whether the student's father has part-time or full-time work (1/0)

* **selfBornUS**: Whether the student was born in the United States of America (1/0)

* **motherBornUS**: Whether the student's mother was born in the United States of America (1/0)

* **fatherBornUS**: Whether the student's father was born in the United States of America (1/0)

* **englishAtHome**: Whether the student speaks English at home (1/0)

* **computerForSchoolwork**: Whether the student has access to a computer for schoolwork (1/0)

* **read30MinsADay**: Whether the student reads for pleasure for 30 minutes/day (1/0)

* **minutesPerWeekEnglish**: The number of minutes per week the student spend in English class

* **studentsInEnglish**: The number of students in this student's English class at school

* **schoolHasLibrary**: Whether this student's school has a library (1/0)

* **publicSchool**: Whether this student attends a public school (1/0)

* **urban**: Whether this student's school is in an urban area (1/0)

* **schoolSize**: The number of students in this student's school

* **readingScore**: The student's reading score, on a 1000-point scale

#### Reading the datasets

```{r}

train <- read.csv("pisa2009train.csv")
test <- read.csv("pisa2009test.csv")
```

#### check the structure of the data

The next step will be to show the structure of variables using the *str*:

```{r}
str(train)
str(test)
```

#### Checking missing values

After checking the structure of the data, we have to check the missing values in each variable using the *summary*.

```{r, results='hide'}
summary(train)
summary(test)
```

* The summary shows there are variables with missing values, therefore, it is better if we remove these missing values.

#### Removing missing values

```{r}
train <- na.omit(train)
test <- na.omit(test)
```

#### Building a Modele using Linear Regression

Before building the linear regression, we have to check the refernce to each factor variable. I this data, the raceeth is a factor variable with seven (7)levels. If we build the linear regresion, R selects the first level alphabetically ("American Indian/Alaska Native") as the level of our factor instead of the most common level ("White").

Therefore, we have to set the reference level of the factor vaiable by using the following.

```{r}
train$raceeth <- relevel(train$raceeth, "White")
test$raceeth <- relevel(test$raceeth, "White")
```

Now we can start to build the model using linear regression. I start by including all the predictor variable. 

```{r}
train_Model_1 <- lm(readingScore~., data=train)
```

```{r}
summary(train_Model_1)
```

Computing SSE and RMSE

```{r}
SSE_1 <- sum(train_Model_1$residuals^2)
RMSE_1 <- sqrt(mean(train_Model_1$residuals^2))

SSE_1
RMSE_1
```

Lets remodel the **train_Model_1** by removing all the non-significant variables.

```{r}
train_Model_2 <- lm(readingScore ~ grade + male + raceeth + expectBachelors
                   + motherBachelors + fatherBachelors + computerForSchoolwork
                   + read30MinsADay + publicSchool + schoolSize, data=train)
summary(train_Model_2)
```

And compute SSE and RMSE

```{r}
SSE_2 <- sum(train_Model_2$residuals^2)
RMSE_2 <- sqrt(mean(train_Model_2$residuals^2))

SSE_2
RMSE_2
```

Though RMSE of **train_Model_2** model is slightly higher that **train_Model_1** model; the **train_Model_2** has small number of predictors which is simple and interpretable model. Therefore, **train_Model_2** is the best model for this data set.

Based on the best selected model (train_Model_2), we can conclude:

Grade of the student, the student`strain_Model_2 bachelor's degree, the student's mother obtained a bachelor's degree, the student's father obtained a bachelor's degree, the student has access to a computer for schoolwork, the student reads for pleasure for 30 minutes/day and number of students in school in positively associated with reading score of the student. 

Being male, American Indian/Alaska Native in race/ethnicity, black in race/ethnicity, hispanic in race/ethinicity and student attends a public school are negatively associated with reading score of the student. 

##### Predicting the test data

And now lets predict using the best model (**train_Model_2**) on the test data.

```{r}
predict_test <- predict(train_Model_2, newdata = test)
summary(predict_test)
```

The minimum and maximum predicted reading score is 356.8 and 634.5. 


And we can compute the **R-squared** and **RMSE** of the predicted reading score.

```{r}
SSE_test <- sum((predict_test - test$readingScore)^2)
SST_test <- sum((mean(train$readingScore) - test$readingScore)^2)

R_squared_test <- 1 - SSE_test/SST_test

RMSE_test <- sqrt(mean((predict_test-test$readingScore)^2))

R_squared_test
RMSE_test
```

